---
title: "Role of Piwi-piRNA pathway in somatic and cancer cells"
author: "__Konstantinos Geles__"
date: "Thu Jun 30 2022, Last Update: `r format(Sys.Date(), '%a %b %d %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 3
  html_notebook: null
editor_options:
  chunk_output_type: console
subtitle: UMG PhD Programme of Molecular and Translational Oncology - Circle XXXIV
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This project contains the scripting part of the Doctoral Dissertation of **Konstantinos Geles** with doi:   

# CHAPTER 1: Role of the PIWI-piRNA pathway in Colorectal Cancer (CRC)  
  
## 1.5 Evaluation of piRNA in RIP-seq

###  piRNA identified inside PIWIL1
For this analysis we have use SPORTS 1.0 to perform all the preprocessing, alignment and quantification steps as shown in the ... GitHub Repository.
Using TMM normalization and transforming to CPM we arrive to a table 

Import Libraries
```{r}
library(vroom)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(scales)
library(ggpmisc)
library(ComplexHeatmap)
library(circlize)
library(limma)
library(ggvenn)
```

Import the voom and get the comparisons
```{r}
voom_IPP <- readRDS("Chapter_1_5/3_IPP_PIWIL1_all_DBs/vfit_x1_voomTMM.rds")

#IPP_C_vs_INPUT
topIPP_C_vs_INPUT <- topTable(voom_IPP, coef = "IPP_C_vs_INPUT",
                         number = nrow(voom_IPP),
                         adjust.method = "fdr",
                         sort.by = "p") %>% 
    as_tibble(rownames = "smallRNA") %>% 
    filter(adj.P.Val< 0.01, logFC > 2)
#IPP_N_vs_INPUT
topIPP_N_vs_INPUT <- topTable(voom_IPP, coef = "IPP_N_vs_INPUT",
                         number = nrow(voom_IPP),
                         adjust.method = "fdr",
                         sort.by = "p") %>% 
    as_tibble(rownames = "smallRNA") %>% 
    filter(adj.P.Val< 0.01, logFC > 2)
#IPP_INPUT_vs_noAb
topIPP_INPUT_vs_noAb <- topTable(voom_IPP, coef = "IPP_INPUT_vs_noAb",
                         number = nrow(voom_IPP),
                         adjust.method = "fdr",
                         sort.by = "p") %>% 
    as_tibble(rownames = "smallRNA") %>% 
    filter(adj.P.Val< 0.01, logFC < (-2))
```

find the enriched
```{r}
enriched_C_IPP <- topIPP_C_vs_INPUT %>% 
    filter(!smallRNA %in% topIPP_INPUT_vs_noAb$smallRNA) %>% 
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" ) 

enriched_C_IPP %>% count(Class)

enriched_N_IPP <- topIPP_N_vs_INPUT %>% 
    filter(!smallRNA %in% topIPP_INPUT_vs_noAb$smallRNA) %>% 
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" )

enriched_N_IPP %>% count(Class)

enriched_IPP_common <- bind_rows("N_ab_LFC" = topIPP_N_vs_INPUT, 
                                 "C_ab_LFC" = topIPP_C_vs_INPUT, .id = "ab") %>% 
    filter(!smallRNA %in% topIPP_INPUT_vs_noAb$smallRNA) %>% 
    select(ab:logFC) %>% 
    pivot_wider(names_from = ab, values_from = logFC) %>% 
    drop_na() %>% 
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" ) 

enriched_IPP_common %>% count(Class)
```

PhD theme for plots
```{r}
wes_cols <- c(wesanderson::wes_palettes$BottleRocket2[2:5])

PhD_theme <-
  list(
    scale_fill_brewer(palette = "Set1"),
    theme_bw() +
      theme(
        panel.border = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 20, colour = "black"),
        axis.line = element_line(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1),
        text = element_text(size = 20),
        axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10),
                                    colour = "black"),
        axis.title.y = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10),
                                    colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
        axis.text.y = element_text(size = 20, colour = "black"),
        plot.title = element_text(hjust = 0.5, colour = "black")
      )
  )
```

make df of union
```{r}
enriched_IPP_union <- bind_rows("IPP_Ab1" = enriched_C_IPP, 
                                 "IPP_Ab2" = enriched_N_IPP, .id = "ab") 
```

plot the enriched
```{r}
plot_enri_un <- enriched_IPP_union %>% 
    ggplot(aes(x = Class,  fill = Class)) +
    geom_bar(stat = "count")+
    stat_count(geom = "text", colour = "black", size = 7.5, 
              aes(label = ..count..), position  = position_fill(vjust = 850)) +
    facet_wrap(facets = "ab", ncol = 2) + 
    xlab("sncRNA Class") + 
    ggtitle("Amount of sncRNAs found enriched in RIP-seq\n for COLO205 cell-line") +
    PhD_theme

tiff(filename = file.path("FIG_21_IPP_1_2_COLO205.tiff"),
     compression = "none", height = 12, width = 16,  units = 'in', res = 600)
plot_enri_un
dev.off()
```


### Heatmap of common piRNAs enriched with both Ab in IPP

import the lcpm 
```{r}
sncRNA_IPP_lCPM <- vroom("Chapter_1_5/3_IPP_PIWIL1_all_DBs/IPP_PIWIL_TMM_logcpm_counts.txt") %>% 
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" ) %>%
    filter(sncRNA %in% enriched_IPP_common$sncRNA, Class == "piRNA") %>% 
    select(-Class, -contains("noAb")) %>% 
    rename_with(.cols = contains("COLO205"), 
                .fn = ~ str_remove(.x, "COLO205_IPP_") %>% 
                    str_replace("C_", "1_") %>% 
                    str_replace("N_", "2_")) %>% 
    column_to_rownames("sncRNA") %>% 
    as.matrix()

```

Scale the matrix for the heatmap
```{r}
# make the matrices for the heatmap -----
sc_mat <- sncRNA_IPP_lCPM %>% t() %>% scale() %>% t()
sc_mat %>% dim()
sc_mat %>% head()
hist(sc_mat)
```

make the heatmap
```{r}
f <- colorRamp2(c(-1, 0 ,1),c("#53868B", "#8B8878", "#FFD700"))

ha_1 <- HeatmapAnnotation(Samples = colnames(sc_mat),
            col = list(Samples =
                           rep(wesanderson::wes_palettes$Cavalcanti1[2:4],
                               each = 3) %>% 
                           set_names(colnames(sc_mat))),
            annotation_legend_param = list(
                
                              Samples = list(
                                  title_gp = gpar( fontsize = 20, fontface = "bold"), 
                                 labels_gp = gpar(fontsize = 20)))
                )

ht_1 <- Heatmap(matrix = sc_mat, #data
        top_annotation = ha_1, #annot
        col = f, #colors data  
        row_title_gp = gpar(fontsize = 20),
        column_title_gp = gpar(fontsize = 20),
        column_names_gp = gpar(fontsize = 20),
        column_names_rot = 45,
        heatmap_legend_param = list(
            title_gp = gpar(fontsize = 15, fontface = "bold"),
            labels_gp = gpar(fontsize = 15)),
        show_row_dend = TRUE,
        show_row_names = FALSE,
        show_column_names = TRUE,
        name = "z-score \nabundance",
        clustering_distance_columns = "spearman",
        clustering_method_columns =  "median",
        clustering_method_rows = "median",
        clustering_distance_rows = "spearman",
        row_dend_reorder = TRUE,
        row_title = "piRNAs",
        column_title = str_glue("Heatmap of {nrow(sc_mat)} common piRNAs\nfound enriched between Ab1 and Ab2"),
)

tiff(filename = file.path("FIG_22_piRNA_common_AB1_AB2_Colo205.tiff"),
     compression = "none", height = 10, width = 14,  units = 'in', res = 600)
ht_1
dev.off()

```


PhD theme for plots
```{r}
wes_cols <- c(wesanderson::wes_palettes$BottleRocket2[2:5])
PhD_theme <-
  list(
    scale_fill_brewer(palette = "Set1"),
    theme_bw() +
      theme(
        panel.border = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = 20, colour = "black"),
        axis.line = element_line(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1),
        text = element_text(size = 20),
        axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10),
                                    colour = "black"),
        axis.title.y = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10),
                                    colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
        axis.text.y = element_text(size = 20, colour = "black"),
        plot.title = element_text(hjust = 0.5, colour = "black")
      )
  )
```

### Methylated and immunoprecipitated piRNA
```{r}
# We import the table made from previous chapter with methylated sncRNAs
dge_contrasts_treat <- vroom("Chap_1_4_dge_contrasts_treat_Enriched_methylate_sncRNAs_Cytr_Nuc.txt")

Treat_cyt <- dge_contrasts_treat %>%  
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" ) %>% 
    filter(comparison == "TrevsUntCyt", Class == "piRNA")

Treat_nuc <- dge_contrasts_treat %>%  
    separate(col = smallRNA, into = c("Class", "sncRNA"),sep = "_match_" ) %>% 
    filter(comparison == "TrevsUntNuc", Class == "piRNA")

#IPP piRNAs by both Antibodies
IPP_piRNA <- enriched_IPP_common %>% 
    filter(Class == "piRNA")


venn_list <- list(`piRNAs Methylated in Cytosol` = Treat_cyt$sncRNA,
          `piRNAs Methylated in Nucleus` = Treat_nuc$sncRNA,
          `piRNAs in RIP-seq` = IPP_piRNA$sncRNA
          )

venn_piRNA <- ggvenn(venn_list, show_percentage = FALSE) +  
    ggtitle("Common piRNAs between Treated Nucleus/Cytosolic and Rip-seq samples")+
    PhD_theme +
    theme_void() +
    coord_fixed()

tiff(filename = file.path("FIG_23_piRNA_common_AB1_AB2_Meth_Cyt_Nuc.tiff"),
     compression = "none", height = 10, width = 14,  units = 'in', res = 600)
venn_piRNA
dev.off() 
```
