---
title: "Role of Piwi-piRNA pathway in somatic and cancer cells"
author: "__Konstantinos Geles__"
date: "Wed Jul 13 2022, Last Update: `r format(Sys.Date(), '%a %b %d %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 3
  html_notebook: null
editor_options:
  chunk_output_type: console
subtitle: UMG PhD Programme of Molecular and Translational Oncology - Circle XXXIV
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This project contains the scripting part of the Doctoral Dissertation of **Konstantinos Geles** with doi:   

# CHAPTER 3: PIWI-piRNA pathway in the in-vitro Mouse Cardiomyocyte differentiation
  
## 3.1.1 DEA for sncRNAs in mouse Cardiomyocytes GSE161081

###  

Import Libraries
```{r}
library(readr)
library(dplyr)
library(edgeR)
library(stringr)
library(purrr)
library(vroom)
library(tidyr)
```

#### Add date of the analysis

We use it as an identifier for the folder and generally the analysis
```{r todate_of_analysis}
todate <- format(Sys.time(), "%d_%b_%Y")
```

#### Make the directory for the results of the DE analysis
```{r make dirs}
my_basename <- file.path("Chapter_3/")  ## INPUT name of the main folder 
my_exp <- "piRNA_GSE161081_cardioMouse" ## INPUT name of the analysis
genome_input <- "GRCh38" ## INPUT genome version here
my_tools <- c("salmon", "featureCounts")
dat_path <- file.path(my_basename, str_glue("DEA_{my_exp}_{genome_input}_{todate}"),
                      my_tools) %>% set_names(my_tools)
dat_path %>% map(~dir.create(., recursive = TRUE))
```

#### 2. Extract normalized objects 

We will work with TMM normalization of voom with quality weights transformed
```{r extract norm dgl}

fc <- vroom("Chapter_3/EDA_mouse_cardiomyocytes_GRCm38_21_Jul_2022/featureCounts/annotation_fc.txt")
fc_vm_QW_TMM <-  read_rds("Chapter_3/EDA_mouse_cardiomyocytes_GRCm38_21_Jul_2022/featureCounts/list_norm_dgls_featureCounts.rds") %>% 
    magrittr::extract2("voomQW_TMM")

salmon_vm_QW_TMM <-  read_rds("Chapter_3/EDA_mouse_cardiomyocytes_GRCm38_21_Jul_2022/salmon/list_norm_dgls_salmon.rds") %>% 
    magrittr::extract2("voomQW_TMM")
```

#### 3. Create the design matrix

If we load the voom object we can extract the design matrix 
```{r design}
design <- salmon_vm_QW_TMM$design
```

#### 4. Limma 
```{r limma_DE}
nc_RNA_categories <- vroom("Chapter_3/EDA_mouse_cardiomyocytes_GRCm38_21_Jul_2022/featureCounts/annotation_fc.txt") %>% 
  select(gene_id = GeneID, gene_type) %>% 
  distinct(gene_id, .keep_all = TRUE) 

## makeContrasts ----
con_mat <- makeContrasts(
  iCM_v_CM = CSC_CM - CM,
  CM_v_CSC = CM - CSC,
  iCM_v_CSC = CSC_CM - CSC,
  levels = design)

## salmon ----
salmon_vm_QW_TMM <- lmFit(salmon_vm_QW_TMM, design = design)
salmon_vm_QW_TMM <- contrasts.fit(salmon_vm_QW_TMM, con_mat)
salmon_vm_QW_TMM <- eBayes(salmon_vm_QW_TMM, robust = TRUE)

salmon_DES_long <- con_mat %>% 
    colnames() %>% 
    set_names() %>% 
 map(~salmon_vm_QW_TMM %>% topTable(., coef = .x,
                             confint = TRUE,
                             number = nrow(.),
                             adjust.method = "fdr",
                             sort.by = "p") %>% 
  as_tibble(rownames = "smallRNA")) %>%
  bind_rows(.id = "contrast") %>% 
  mutate(quantification = "salmon", .after = smallRNA) %>% 
  left_join(nc_RNA_categories, by = c("smallRNA" = "gene_id"))

## for venn ---
salmon_vm_TMM_p <- salmon_DES_long %>% 
    filter(contrast == "iCM_v_CM") %>% 
    mutate(salmon_voomQW = 
           case_when(
             select(., starts_with("adj.P.Val")) >= 0.05 ~ 0L,
             select(., starts_with("logFC")) > 0 ~ 1L,
             select(., starts_with("logFC")) < 0 ~ -1L
           )) %>% 
  select(smallRNA , salmon_voomQW) 

## featureCounts ----
fc_vm_QW_TMM <- lmFit(fc_vm_QW_TMM, design = design)
fc_vm_QW_TMM <- contrasts.fit(fc_vm_QW_TMM, con_mat)
fc_vm_QW_TMM <- eBayes(fc_vm_QW_TMM, robust = TRUE)

fc_DES_long <- con_mat %>% 
    colnames() %>% 
    set_names() %>% 
    map(~fc_vm_QW_TMM %>% topTable(., coef = .x,
                             confint = TRUE,
                             number = nrow(.),
                             adjust.method = "fdr",
                             sort.by = "p") %>% 
  as_tibble(rownames = "smallRNA")) %>%
  bind_rows(.id = "contrast") %>%
  mutate(quantification = "featureCounts", .after = smallRNA) 

## for venn ---
fc_vm_TMM_p <- fc_DES_long %>% 
  filter(contrast == "iCM_v_CM") %>% 
  mutate(fc_voomQW = 
           case_when(
             select(., starts_with("adj.P.Val")) >= 0.05 ~ 0L,
             select(., starts_with("logFC")) > 0 ~ 1L,
             select(., starts_with("logFC")) < 0 ~ -1L
           ))%>% 
  select(smallRNA , fc_voomQW )

## venn diagram for salmon/fc limma -----
results <-  salmon_vm_TMM_p %>% 
  inner_join(fc_vm_TMM_p) %>%  
    select(-smallRNA)

pdf(file.path(dirname(dat_path[1]), str_c("venn_diagram_DE_salmon_fC_limma_iCM_v_CM_", colnames(con_mat)[1],".pdf")))
vennDiagram(results, 
    include=c("up", "down"),
    counts.col=c("red", "blue"),
    circle.col = c("red", "blue", "green3"))
dev.off()

## join both results ----
identical(fc_DES_long %>% names(), salmon_DES_long %>% names())


## sncRNA names
nc_RNA_names <- file.path("sncRNA_piRNBnk_RNACent_gene_names_GRCm38_v34.gtf.gz") %>% 
  plyranges::read_gff2() %>% 
  as_tibble() %>% 
  select(smallRNA = gene_id, external_id, sncRNA_name) %>% 
  distinct(smallRNA, .keep_all = TRUE) 

## long formats
GSE161081_cardioMouse_all_comp_long_format <- bind_rows(salmon_DES_long, fc_DES_long ) %>% 
    left_join(nc_RNA_names)

GSE161081_cardioMouse_all_comp_long_format %>% vroom_write(file.path(dirname(dat_path[1]), 
                        str_c("GSE161081_cardioMouse_all_comparisons_long_voom_TMMQW_salmon_fc_LFCs_", 
                              todate,".txt")))
```

