---
title: "Role of Piwi-piRNA pathway in somatic and cancer cells"
author: "__Konstantinos Geles__"
date: "Wed Jul 13 2022, Last Update: `r format(Sys.Date(), '%a %b %d %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 3
  html_notebook: null
editor_options:
  chunk_output_type: console
subtitle: UMG PhD Programme of Molecular and Translational Oncology - Circle XXXIV
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This project contains the scripting part of the Doctoral Dissertation of **Konstantinos Geles** with doi:   

# CHAPTER 2: Data Analysis Workflow for small-RNAseq focused on piRNAs  

## 2.2.3 c) GSE160432 dataset comparison between SPAR and WIND results

Here we compare the  results we got from SPAR and WIND analysis of the public dataset [GSE160432](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE160432).

### import libraries
```{r}
library('dplyr') 
library('tibble')          
library('stringr')
library('tidyr')
library('vroom')
suppressPackageStartupMessages({
  
  library('plyranges')
  library('tximport')
  library('edgeR')
  library('NOISeq')
  library('rafalib')
  library('pheatmap')
  library('RColorBrewer')
})
```

### import the WIND results and check lFC of two miRNAs
```{r}
DE_WIND <- vroom("Chapter_2_2/WIND/DEA_piRNA_GSE160432_CRC_GRCh38_20_Jul_2022/all_comparisons_long_voom_TMMQW_salmon_fc_LFCs_20_Jul_2022.txt")

# check the two specific miRNAs that the article is focusing for WIND results
DE_WIND %>%
  filter(str_detect(external_id, "-1246|-215-5p")) %>%
  pivot_wider(names_from = quantification, values_from = c(logFC,P.Value, adj.P.Val)) %>%
  select(-c(smallRNA, CI.L:B, Chr:seq_RNA))
```

####> A tibble: 6 x 10
  contrast      gene_type external_id    sncRNA_name    logFC_salmon logFC_featureCounts P.Value_salmon P.Value_featureCounts adj.P.Val_salmon adj.P.Val_featureCounts
  <chr>         <chr>     <chr>          <chr>                 <dbl>               <dbl>          <dbl>                 <dbl>            <dbl>                   <dbl>
1 Tumour_v_Ctrl miRNA     hsa-miR-1246   hsa-mir-1246         3.78                 NA          8.25e-10             NA            0.0000000379               NA       
2 Tumour_v_poly miRNA     hsa-miR-1246   hsa-mir-1246         3.69                 NA          3.89e- 7             NA            0.0000159                  NA       
3 poly_v_Ctrl   miRNA     hsa-miR-1246   hsa-mir-1246         0.0842               NA          9.09e- 1             NA            0.953                      NA       
4 Tumour_v_Ctrl miRNA     hsa-miR-215-5p hsa-mir-215-5p      NA                    -3.22      NA                     4.77e-15    NA                           1.70e-12
5 Tumour_v_poly miRNA     hsa-miR-215-5p hsa-mir-215-5p      NA                    -1.87      NA                     2.36e- 6    NA                           3.37e- 5
6 poly_v_Ctrl   miRNA     hsa-miR-215-5p hsa-mir-215-5p      NA                    -1.35      NA                     8.90e- 5    NA                           7.78e- 4


### import the SPAR results and check lFC of two miRNAs
```{r}
DE_SPAR <- vroom("Chapter_2_2/SPAR/DEA_CRC_GRCh38_23_Jun_2022/DE_SPAR_vm_QW_TMM.txt")

# check the two specific miRNAs that the article is focusing for SPAR results
DE_SPAR %>%
  filter(str_detect(smallRNA, "-1246|-215-5p")) %>%
  select(-c( CI.L:P.Value, B))
```

####> A tibble: 3 x 5
  contrast     smallRNA                                  logFC adj.P.Val GeneClass
  <chr>        <chr>                                     <dbl>     <dbl> <chr>    
1 CRC_v_ctrl   chr1:220117915:220117936:-:hsa-miR-215-5p -3.55  3.85e-14 mir-5p   
2 CRC_v_Polyp  chr1:220117915:220117936:-:hsa-miR-215-5p -1.92  9.00e- 5 mir-5p   
3 Polyp_v_ctrl chr1:220117915:220117936:-:hsa-miR-215-5p -1.63  2.09e- 5 mir-5p 

The miR-1246 is not identified in SPAR results

we will check if at the same position with the newer version can be found DE
```{r}
# check if we can find the same chromosome and start for another possible mirna name

DE_SPAR %>%
  separate("smallRNA", c("chr","start","end","strand","smallRNA","DQ"), sep = ":") %>%
  filter(chr == "chr2", str_detect(GeneClass, "mir")) %>% 
  arrange(start) # nope

# check if we can find the same chromosome and start for another sncRNA name
DE_SPAR %>%
  separate("smallRNA", c("chr","start","end","strand","smallRNA","DQ"), sep = ":") %>%
  filter(chr == "chr2", between(start,176600970, 176601052)) %>% 
  arrange(start)# nope

# check if we can find the same chromosome and start for another sncRNA name in initial counts
initial_df <- readr::read_rds("Chapter_2_2/SPAR/EDA_CRC_GRCh38_23_Jun_2022_SPAR/initial_df.rds")
initial_df %>%
  separate("smallRNA", c("chr","start","end","strand","smallRNA","DQ"), sep = ":") %>%
  filter(str_detect(smallRNA, "-1246")) %>% 
  select(1:10)
```
#### A tibble: 2 × 10
  chr   start     end       strand smallRNA     DQ    GeneClass    SRR12936664 SRR12936665 SRR12936666
  <chr> <chr>     <chr>     <chr>  <chr>        <chr> <chr>              <dbl>       <dbl>       <dbl>
1 chr2  176600979 176601052 -      hsa-mir-1246 NA    miRNAprimary           0           0           0
2 chr2  176601023 176601042 -      hsa-miR-1246 NA    mir-5p3pno             0           0           0

From this result we can see that the hsa-mir-1246 is not found expressed. 


### How many miRNA and piRNA are DE with SPAR
```{r}
DE_SPAR_CRCvCTRL <- DE_SPAR %>% 
  filter(contrast == "CRC_v_ctrl", adj.P.Val < 0.05) %>% 
  separate("smallRNA", c("chr","start","end","strand","smallRNA","DQ"), sep = ":") 

# how many miRNA are DE? 
DE_SPAR_CRCvCTRL %>% 
  filter(str_detect(GeneClass, "mir|miR")) %>%
  count(GeneClass, sort = TRUE) # 147

# how many piRNA are DE? 
DE_SPAR_CRCvCTRL %>% 
  filter(str_detect(GeneClass, "pi")) %>%
  count(smallRNA, sort = TRUE) # 19
```

### How many miRNA and piRNA are DE with WIND
```{r}
DE_WIND_CRCvCTRL <- DE_WIND %>% 
  filter(contrast == "Tumour_v_Ctrl", adj.P.Val < 0.05) %>% 
  select(-c( CI.L:P.Value, B, Chr:seq_RNA))

# how many miRNA are DE? 
DE_WIND_CRCvCTRL %>% 
  filter(str_detect(gene_type, "precursor|miRNA")) %>%
  pivot_wider(names_from = quantification, 
              values_from = c(logFC, adj.P.Val)) #700

DE_WIND_CRCvCTRL %>% 
  filter(str_detect(gene_type, "precursor|miRNA")) %>% 
    count(quantification) # FC 616, salmon 284

# how many piRNA are DE? 
DE_WIND_CRCvCTRL %>% 
  filter(str_detect(gene_type, "pi")) %>%
  count(quantification, sort = TRUE) # FC 73, salmon 48
```

### Compare the DE miRNA found in WIND and SPAR in CRC vs CTRL
```{r}
 DE_SPAR_mirna <- DE_SPAR_CRCvCTRL %>%
  filter(str_detect(GeneClass, "mir|miR")) %>%
  mutate(smallRNA = tolower(smallRNA)) %>%
  count(smallRNA ) %>% # 146 miRNA
  pull(smallRNA)

DE_WIND_CRCvCTRL %>%
  filter( gene_type %in% c("precursor_RNA", "miRNA")) %>%
  pivot_wider(names_from = quantification, values_from = c(logFC,adj.P.Val)) %>% #700
  count(external_id) #external_id:694 #sncRNA_name:688

DE_WIND_CRCvCTRL %>%
  filter(gene_type %in% c("precursor_RNA", "miRNA")) %>%
  mutate(external_id = tolower(external_id)) %>%
  filter(str_detect(sncRNA_name, str_c(DE_SPAR_mirna, collapse = "|"))) %>%
  pivot_wider(names_from = quantification, values_from = c(logFC,adj.P.Val)) %>% 
  count(sncRNA_name, sort = T) # 100 common names

# make the mirna WIND table to join
mirna_wind_join <- DE_WIND_CRCvCTRL %>%
  filter(gene_type %in% c("precursor_RNA", "miRNA")) %>%
  mutate(external_id = tolower(external_id)) %>%
  filter(str_detect(sncRNA_name, str_c(DE_SPAR_mirna, collapse = "|"))) %>%
  pivot_wider(names_from = quantification, values_from = c(logFC,adj.P.Val)) %>%
  rename_with(.fn = ~str_c("wind_", .), .cols = c(-contrast)) 

# the mirna SPAR table to join

mirna_SPAR_join <- DE_SPAR_CRCvCTRL %>%
  filter(str_detect(GeneClass, "mir|miR")) %>%
  unite(chr:strand, col = "GenRange", sep = "_") %>%
  mutate(smallRNA = tolower(smallRNA)) %>%
  distinct(smallRNA, .keep_all = TRUE) %>%
  select(-c(contrast, DQ, CI.L:P.Value, B)) %>%
  rename_with(.fn = ~str_c("SPAR_", .)) 

# make a common table
mirna_wind_join %>% 
  inner_join(mirna_SPAR_join, by = c("wind_sncRNA_name" = "SPAR_smallRNA")) %>%
  vroom_write("Chapter_2_2/DE_miRNA_common_SPAR_WIND_vm_QW_TMM.txt")
```
